Scaling up the hyperparameter grid for Naomi

### 2023-07-05 src/dev_scale-grid/scale-grid.Rmd
Extend and simplify scale grid development, including first try at dealing with scaling #58

  **Task** We would like to begin to integrate over the hyperparameters using multiple nodes. The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical. In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r cov, class.source = 'fold-hide', fig.cap="Covariance matrix."}
C <- sd_out$cov.fixed
C_df <- reshape2::melt(as.matrix(C))

C_plot <- C_df %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

C_plot
```

@@

```{r}
quad$modesandhessians[, 1:24] %>%
  pivot_longer(cols = everything()) %>%
  mutate(id = rep(1:24, times = 2)) %>%
  ggplot(aes(x = reorder(name, id), y = value)) +
    geom_point(alpha = 0.8) +
    coord_flip() +
    labs(x = "", y = "Value") + 
    theme_minimal()
```

@@

```{r scree, class.source = 'fold-hide', fig.cap="Scree plot."}
tv_df <- data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
)

ggplot(tv_df, aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()

ggsave("total-variation.png", h = 3, w = 6.25)
```

Based on Figure \@ref(fig:scree) with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.

@@

```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
rownames(E) <- hypers

reshape2::melt(as.matrix(E)) %>%
  ggplot(aes(x = Var1, y = factor(Var2), fill = value)) +
  labs(x = "", y = "", fill = "") +
  geom_tile() +
  coord_flip() +
  scale_fill_gradientn(
    colours = c("#E69F00", "white", "#009E73"),
    rescaler = ~ scales::rescale_mid(.x, mid = 0),
    limits = c(-1.1, 1.1)
  ) +
  labs(x = "Hyper", y = "Principal component loading") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

@@

C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
  ggplot(aes(x = as.factor(Var1), y = as.factor(Var2), fill = value)) +
  labs(x = "", y = "", fill = "C_s[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "none"
  )

{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")}

@@

Put `logit` and `log` on parity somehow!

>>>>

  **Task** We would like to begin to integrate over the hyperparameters using multiple nodes. The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical. In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using different varieties of principal components analysis.
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r cov, class.source = 'fold-hide', fig.cap="Covariance matrix."}
C <- sd_out$cov.fixed
(C_plot <- plot_matrix(C))
```

@@

```{r}
quad$modesandhessians[, 1:24] %>%
  pivot_longer(cols = everything()) %>%
  mutate(id = rep(1:24, times = 2)) %>%
  ggplot(aes(x = reorder(name, id), y = value)) +
    geom_jitter(shape = 1, height = 0, width = 0.2) +
    coord_flip() +
    labs(x = "", y = "Value") + 
    theme_minimal()
```

@@

```{r scree, class.source = 'fold-hide', fig.cap="Scree plot."}
plot_total_variation(eigenC, label_x = 20)

ggsave("total-variation.png", h = 3, w = 6.25)

tv <- cumsum(eigenC$values) / sum(eigenC$values)
```

Based on Figure \@ref(fig:scree) with 5 dimensions included, we can explain `r signif(100 * tv[5], 3)`% of the total variation.
Or, with 10 dimensions included, that percentage increases to `r signif(100 * tv[10], 3)`%.

@@

```{r pc-loadings, class.source = 'fold-hide', fig.cap="PC loadings."}
rownames(E) <- hypers
plot_pc_loadings(eigenC)

@@

C_s_plot <- plot_matrix(C_s)
{C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + theme(legend.position = "none") + labs(subtitle = "Reduced rank")}

@@

Eventually we want to put parameters on the log and logit scales in parity.

## Using the correlation matrix

```{r}
C <- sd_out$cov.fixed
d <- 1 / (sqrt(diag(C)))
R <- diag(d) %*% C %*% diag(d)
plot_matrix(R)
```

```{r}
eigenR <- eigen(R)
plot_total_variation(eigenR, label_x = 5)
```

```{r}
rownames(eigenR$vectors) <- hypers
plot_pc_loadings(eigenR)
```

## Scale standardised system

```{r}
C <- sd_out$cov.fixed
var <- diag(C)

df_std <- data.frame(var = var) %>%
  tibble::rownames_to_column("par") %>%
  mutate(
    scale = fct_recode(str_extract(par, "^[^_]+"), "Log" = "log", "Logit" = "logit", "Other" = "OmegaT")
  ) %>%
  group_by(scale) %>%
  mutate(mean_var = mean(var)) %>%
  ungroup() %>%
  mutate(var_std = var / mean_var)

df_std %>%  
  pivot_longer(
    cols = c("var", "var_std"),
    names_to = "method",
    values_to = "value"
  ) %>%
  mutate(
    method = fct_recode(method, "Default" = "var", "Scale standardised" = "var_std"),
    start = str_extract(par, "^[^_]+"),
    scale = fct_recode(start, "Log" = "log", "Logit" = "logit", "Other" = "OmegaT") 
  ) %>%
  ggplot(aes(x = reorder(par, value), y = value, col = scale)) +
    geom_point() +
    facet_grid(~ method) +
    coord_flip() +
    scale_color_manual(values = c("#56B4E9","#009E73", "#E69F00")) +
    labs(x = "Hyperparameter", y = "Marginal variance", col = "Scale") +
    theme_minimal()

Rs <- diag(1 / sqrt(df_std$mean_var)) %*% C %*% diag(1 / sqrt(df_std$mean_var))
plot_matrix(Rs)

eigenRs <- eigen(Rs)
plot_total_variation(eigenRs, label_x = 5)

rownames(eigenRs$vectors) <- hypers
plot_pc_loadings(eigenRs)
```

# Testing pruning

Remove any nodes which have small enough weight.
